---
title: "brain"
output: html_document
date: "2025-04-28"
---

## **QUESTION:** 

## What age range receives the largest number of traumatic brain injuries?

## 

### **THE DATA:**

What data did you find to help answer that question?

-   We found CDC data that grouped brain injury total and rate per 100,000 estimates based on injury mechanism, age, type of measure (emergency visit, post-mortem examination, hospitalization) and year estimated

    -   This will allow us to see what ages and what activities have the highest likelihoods of getting a TBI

### **RELEVANT BACKGROUND:**

-   The CDC estimates totals nationwide based on the large quantity of data they have, so our numbers are not technically true, they are just really good guesses.

-   The number of TBIs that america sees each year is in the hundreds of thousands

-   There are three different ways that TBIs are reported

    -   Hospitalization - anytime someone would be hospitalized temporarily by a TBI

    -   Emergency Care - anytime a TBI would result in being brought to an emergency care facility (ER, urgent care, …)

    -   Death - anytime a TBI is reported post-mortem

### **Basic EDA:**

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

age_link <- ".\\tbi_age.csv"
military_link <- ".\\tbi_military.csv"
year_link <- ".\\tbi_year.csv"
tbi_age <-  read.csv(age_link)
tbi_military <-  read.csv(military_link)
tbi_year <-  read.csv(year_link)
tbi_age %>% 
  distinct(age_group)
tbi_year %>%
  distinct(year)
tbi_age %>% 
  distinct(injury_mechanism)
tbi_age %>% 
  distinct(type)

```

-   tbi_age:

    -   Five columns: age_group, type, injury_mechanism, number_est, rate_est

    -   age_group: Categorizes ages 0-75+ into 9 bins of 10 years each

    -   type: The result of the injury

        -   Emergency Department Visit, Hospitalization, or Death

    -   injury_mechanism: How the person got injured

        -   Assault, Intentional Self-harm, Motor Vehicle Crashes, Other or no mechanism specified, Other unintentional injury (mechanism unspecified, Unintentional Falls, Unintentionally Struck By or Against an Object

    -   number_est: Total number of TBI patients in that category

    -   rate_est: Rate of TBI of that category per 100,000

-   tbi_year: 

    -   Five Columns: injury_mechanism, type, year, rate_est, number_est

    -   Same definition for injury_mechamism, type, rate_est, and number_est as tbi_age

    -   Year: Year the rate/number of patients was from spanning from every year between 2009 and 2014

-   We chose not to use the military data and focus on just civilians so that we can more validly compare the age data and year data

```{r}

tbi_age %>% 
  filter(age_group != "0-17", age_group !="Total") %>% 
  mutate(age_group = factor(age_group, levels = c("0-4", "5-14", "15-24", "25-34", 
                                                  "35-44", "45-54", "55-64", 
                                                  "65-74", "75+"))) %>% 
  ggplot(aes(age_group, rate_est)) +
  geom_col() +
  facet_grid(injury_mechanism ~ type) +
theme(
  strip.text.x = element_text(angle = 0, hjust = 0.5),
  strip.text.y = element_text(angle = 0, hjust = 1),
  axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
)

tbi_year %>% 
  filter(injury_mechanism != "Total") %>%
  ggplot(aes(year,rate_est, color = injury_mechanism)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  scale_y_log10() +
  facet_wrap(~type) +
  geom_line() 

```

### Observations: 

It seems that the highest rates come from injury mechanisms of unintentional falls and other blows to the head and motor vehicle crashes. In particular, the motor vehicle rates are high for people age 18-36, and unintentional fall rates are high for young and elderly people.

### Quantified Uncertainty:

-   We feel that it does not make sense for our data to come up with uncertainty ourselves, as we our working with quantified totals, which are already estimates themselves.

-   Looking at the CDC data, we see that, before they updated their concussion research program in 2019, they quantified that they only captured the data of 1 in 9 concussions

    -   <https://www.cdc.gov/traumatic-brain-injury/programs/index.html>

    -   This is quite significant, but also, at the scale of hundreds of thousands (about 625,000), makes sense, and the trends you see from 1/9 of this number most likely are quite similar. We cannot deny however, that there may be a non-report bias somehow - that there is a chance that certain cases are more likely to be reported, which would skew the data.

    -   We personally feel that the CDC is a reputable source for good quality data, and are comfortable with the conclusions we make despite this mild uncertainty.

# Conclusion:

It seems that, across age groups, people aged 75+ and 0-4 are the most at risk of getting a TBI, and through injury mechanism data, motor vehicle drivers and people suffering from falls are the most at risk of getting a TBI among all the things people are doing. It makes sense that the highest proportion of traumatic brain injuries were for individuals 75+ years in age and from falling. This is because as people age they lose balance and don't realize the rapid rate at which they can no longer control aspects of their body and movement.

### Remaining Questions:

-   Is there a non-report bias?

-   Are certain groups of people not shown in our data more likely to be affected by TBIs (thinking about ethnicity, gender, …)?
